---
title: "Iris Clustering"
output: html_notebook
---

I have implemented my own k-means clusering algorithm in `k_means_clusering.R`, which is imported here. I will use that to explore the iris dataset, which can be imported or called from base R directly.

```{r}
library(tidyverse)
source("k_means_clustering.R")
```

I will import the .csv also provided in this directory, and visualise it:

```{r}
df_iris <- read.csv(file = "iris.csv")  # Could also use built-in dataset `iris`

# png(filename = "images/iris_dataset.png")
pairs(df_iris[1:4], main="Iris Data (red = setosa, green = versicolor, blue = virginica)",
      pch=20, bg=c("red","green","blue")[unclass(df_iris$Species)])
# dev.off()
```

First remove the categories from the original dataset.

```{r}
df_x <- df_iris %>% 
  select(-Species)  
```

Then determine cluster IDs by running the k-means clustering algorithm:

```{r}
clusters <- run_kmeans(df_x, 3)
```

We will now compare those predicted clusters to the species labels.

```{r}
df <- df_iris %>% 
  as.data.frame() %>% 
  mutate(cluster=clusters)
table(df$Species, df$cluster)
```

We see that our algorithm did a good job, and was able to differentiate between species correctly in most cases. The overlap between _versicolor_ and _virginica_ suggest these two species are more similar to each other than they are to _setosa_.

But this was helped by knowing that there were three species present! If we chose k = 2 instead:

```{r}
clusters <- run_kmeans(df_x, 2)
df <- df_iris %>% 
  as.data.frame() %>% 
  mutate(cluster=clusters)
table(df$Species, df$cluster)
```

We see that the algorithm this time simply differentiated between _setosa_ and the other species - a logical output given that this seems to be better separated from the other species based on the previous result.