---
title: "Using Statistical Distributions"
output: html_notebook
---

# Basic Evaluations

<br>$P(X = 112)$ where $X$ is binomial with $n = 200, \ p = 0.6$.


```{r}
n <- 200
p <- 0.6
x <- 112

value <- choose(n, x) * (p ^ x) * ((1 - p)^(n - x)) 
print(paste("My func: ", value))
print(paste("Built-in: ",dbinom(x, size = n, prob = p)))
```

<br>$P(X ≥ 4)$ where $X$ is Poisson with parameter $\lambda = 8$

```{r}
lambda <- 8
x <- 4

poisson_pdf <- function(rate, value){
  return(exp(-rate) * (rate ^ value) / factorial(value))  
}
poisson_cdf <- function(rate, value){
  sum <- 0
  for (x in 0:value){
    sum = sum + poisson_pdf(rate, x)
  }
  return(sum)
}

value <- (1 - poisson_cdf(lambda, x))
print(paste("My func: ", value))
print(paste("Built-in: ",(1 - ppois(x, lambda))))
```

<br>$P(1 < X < 2)$ where $X$ is Exponential with parameter $\lambda = 2$

```{r}
lambda <- 2
lower_val <- 1
upper_val <- 2

exp_cdf <- function(value, lambda){
  return(1 - exp(-lambda * value))
}
print(paste("My func: ",(exp_cdf(upper_val, lambda) - exp_cdf(lower_val, lambda))))
print(paste("Built-in: ",(pexp(upper_val, lambda) - pexp(lower_val, lambda))))
```

<br>$P(X < 2)$ where $X$ is normal with $\mu = 3, \sigma = 7$

```{r}
print(paste("P(X < 2) =", pnorm(2, 3, 7)))
```

# Exponential Model Distribution

$$ f(x;\lambda) = \lambda e^{-\lambda x} \qquad x \ge 0 $$

I simulate $1000$ groups of $k$ copies of an exponential random variable
with parameter $\lambda = 2$, where $k = 1, 2, 10, 100$.

```{r}
# Simulation

sim_exp_rand_groups <- function(lambda, group_size, n_group){
  # Simulates n_groups of random variables, from an exponential
  # distribution with parameter lambda
  for(i in 1:n_group){
    row <- rexp(group_size, lambda)
    if(i==1){
      array <- row
      next
    }
    array <- rbind(array, row)
  }
  return(array)
}

```


For each k do the following, I compute the mean of each group, and use a normal Q–Q plot to compare it to a normal distribution. I also plot a histogram, superimposing a normal curve with the same mean and variance.


```{r}
# Plotting

library(ggplot2)

norm_qq_plot <- function(data){
  qqnorm(data, frame = FALSE)
  qqline(data, col = "steelblue", lwd = 2)
}

norm_hist_plot <- function(data, lambda, k){
  hist(data, freq=FALSE, xlab = "Mean of Exponential Random Variable Groups")
  curve(dnorm(x, mean=(1 / lambda), sd=(1 / (lambda * sqrt(k)))), 
      col="darkblue", lwd=2, add=TRUE, yaxt="n")
}

```

```{r}
# Running 

main <-  function(lambda, group_size, n_group){
  data <- sim_exp_rand_groups(lambda, group_size, n_group)
  means <- rowMeans(data)
  norm_qq_plot(means)
  norm_hist_plot(means, lambda, group_size)
}


n_group <- 1000
k <- 100  # Can vary this to see the effect of sample size
lambda <- 2
main(lambda, k, n_group)
```
Note the good agreement with the normal distribution when k is large enough, whereas the prevalance of extreme values in the smaller group sizes resulting in deviation from the normal distribution.
