---
title: "Genomic Statistics"
output: html_notebook
---

# Descriptive Statistics

The following analysis uses data from the completely sequenced genome of E. coli,
more precisely strain K-12, substrain MG 1655, version M52 short. 
Methods used are adapated from a lab written by Terry Speed and Bin Yu.

```{r}
# Read in data from local file

# Takes a text file listing a genome sequence with one element per line
ecp <- scan(file = "Ecoli.txt", what=character())

```

Consider the frequency of each base in this sequence:

```{r}
print(round((table(ecp) / length(ecp)), 4))
```

We may also consider the contingency table for all pairs of bases within the sequence:

```{r}
L <- length(ecp)
print(round(table(ecp[-L],ecp[-1]) / L, 3))
```

This suggests the probability of observing a given base (such as G) is not independent of the value of the previous base.

# Purine Counts

Divide up the data into blocks of multiple base pairs (bp).
Count the number of purines (i.e. A or G). 

```{r}
library(matrixStats)

split_data <- function(vec_data, row_len){
  row_num <- floor(length(vec_data) / row_len)
  return(matrix(vec_data, nrow = row_num, byrow = TRUE))
}

count_purines <-  function(vec_data, row_len){
  # Groups bases into subsets of size row_len
  # Counts number of purines and plots histograms
  mat_data <- split_data(vec_data, row_len)
  count_A <- rowCounts(mat_data, value = "A")
  count_G <- rowCounts(mat_data, value = "G")
  count_tot <- count_A + count_G
  return(count_tot)
}

```

We may now plot this data in comparison to a normal distribution: 

```{r}
library(fitdistrplus)

counts <- count_purines(ecp, 10000)
fit <- fitdist(counts, "norm")
denscomp(fit, xlab = "Number of purines in group")
```

We observe here that smaller group sizes result in a better fit to the normal distribution, as we have better smpling for more groups.
